% 	This template is  MIT licensed.

% 	Basic file to demonstrate the usage of this LaTeX template.
% 	You can build your own paper/thesis on top of this file.
% 	Simply adjust the document class and all metadata and start working.

\documentclass[
	language=english, % set to english or german
	type=master, % set to bachelor, master or seminar
]{isthesis}

% Graphics rendering using TikZ
% See: https://en.wikibooks.org/wiki/LaTeX/PGF/TikZ
\usepackage{tikz}
% Include required TikZ libraries here, some exemplary libraries are pre-included
\usetikzlibrary{calc}
\usetikzlibrary{matrix}
\usetikzlibrary{positioning}
\usetikzlibrary{shapes.geometric}

%Add your library here
\addbibresource{library.bib}
% \usepackage{newtxtext,newtxmath}


% Import acronyms
\input{config/acronyms}

% Import custom commands
\input{config/commands}

% Document meta information
\isthesis{
    title={Neuro-SAM},
    sub-title={Foundation Models for Dendrite and Dendritic Spine Segmentation},
    author-name={Nipun Arora}, % Separate multiple authors with commas
    author-matriculation={23084364},
    author-address={Nürnberger Straße 74},
    author-zip={91052},
    author-city={Erlangen},
    principal-supervisor={Prof. Dr. Andreas Kist}, % This must be a professor
    % associate-supervisor={Firstname Lastname}, % This is your main supervisor, i.e., a post doc or doctoral student
    % medical-supervisor={Firstname Lastname},% If required, define your medical supervisor here
    group={Juniorprofessur for Artificial Intelligence in Communication Disorders},
    group-institute={Department Artificial Intelligence in Biomedical Engineering, \\ Friedrich-Alexander-Universität Erlangen-Nürnberg},
    % studies={M.Sc. Medical Engineering},
    %associate-group={}, % When the thesis is done in cooperation with another chair, add it here
    %associate-group-institute={}, % add cooperating institute or university here
    submission-date={2025-09-01} % The date you handed in your document: Format yyyy-mm-dd
    %primary-logo={}, % Uses the FAU logo by default
    %primary-logo-height={}, % Uses 16mm as default height
    %secondary-logo={}, % Logo of the secondary institution (cooperating chair/university), USES Faculty logo by default
    %secondary-logo-height={} % Uses 16mm as default height
}

\begin{document}
    % Title page
    \newcounter{savepage}
    \maketitle

    % Acknowledgements
    \begin{danksagung}
    I would like to extend my gratitude to my supervisor, Prof. Dr. Andreas Kist, for his invaluable guidance, patience, and encouragement throughout the course of my thesis. His expertise in the domain and constructive feedback were essential in shaping the direction of my work. I would also like to thank the experts behind DeepD3 \cite{Fernholz_2024} who laid the foundation and open-sourced the datasets for further developments. A special thanks to Sophie and René for proofreading my thesis and providing necessary feedback. I also wish to extend my heartfelt appreciation to everyone at Anki Lab for giving valuable insights when required.
    \end{danksagung}

    \begin{abstract}
	    % Add your abstract here:
    The quantification of dendrites and dendritic spines is essential to understand synaptic connectivity, plasticity, and circuit-level computations in the brain. Despite advances in high-resolution microscopy, segmentation of these fine structures remains a major bottleneck due to their morphological variability, imaging artifacts, and limitations of existing automated methods. Classical computer vision and deep learning approaches such as DeepD3 have advanced the field, but they often require extensive task-specific tuning and struggle to generalize across modalities and experimental conditions.

    In this thesis, we investigate the potential of vision foundation models, particularly the \gls{SAM} and its variants, for robust and generalizable dendrite and dendritic spine segmentation. We systematically evaluate \gls{SAM}, \gls{SAM}+\gls{LoRA}, and \gls{SAMv2}, analyzing their suitability for fine-scale neural morphology. Building on these insights, we introduce \textbf{Neuro-\gls{SAM}}, an interactive, modular framework that integrates four key components: dendrite path tracing, dendrite segmentation, spine detection, and spine segmentation. Neuro-\gls{SAM} leverages \gls{SAMv2}’s memory-based architecture alongside custom prompt strategies and visualization tools to deliver precise segmentation results with minimal manual input.

    Experimental validation against the DeepD3 benchmark dataset and additional microscopy data demonstrates that Neuro-\gls{SAM} achieves competitive segmentation performance while maintaining strong generalization across species, brain regions, and imaging modalities. The framework bridges state-of-the-art vision models with practical neuroscience workflows, offering a scalable and adaptable solution for interactive analysis of dendritic structures.
    
	\end{abstract}

    % Abstract in german
	% \begin{zusammenfassung}
	%    \lipsum[1]
	% \end{zusammenfassung}
    
    % Table of contents
    \tableofcontents
    
    % List of abbreviations (if you use acronyms)
    \listofabbreviations
    \listoffigures
    
    % Add an empty page (if needed!), otherwise just comment it out
    % \newpage\null\thispagestyle{empty}\newpage

	% storing the last pagenumber
    \setcounter{savepage}{\value{page}}
    
    % Content
    \begin{content}
        % Add your content files:
        % \input{sections/01_TemplateFeatures}
        \input{sections/01_Introduction}
        \input{sections/02_BackgroundRelatedWork}
        \input{sections/03_Methodologies}
        \input{sections/04_Results}
        \input{sections/05_Discussion}
        \input{sections/06_Conclusion}
    \end{content}
    
    \pagenumbering{Roman}
    \setcounter{page}{\numexpr\value{savepage}}

    % References
    \references{}
    % \printbibliography
    
    % Appendix
     \begin{appendix}
        % In the appendices, use \section{} instead of \chapter{}
         \input{appendices/appendix}
     \end{appendix}


    % Declaration of authorship
    % \authorshipstatement[pagenumbering=false]
    \authorshipstatement[pagenumbering=true]
    % \authorshipstatement[pagenumbering=only]
    
    % Consent form for use of plagiarism detection software
    % Not yet required
    % \consentform[pagenumbering=false]
    % \consentform[pagenumbering=true]
    % \consentform[pagenumbering=only]
    
    % Bonus: Wordcount
    % cd %FOLDER WHERE THE .tex FILES ARE IN %
    % clear
    % texcount -total -q -col -sum *.tex
    
\end{document}